---
title: "Chapter 1 Introduction to R for Bioinformatics"
author: "Sasha Ajay Malkani"
date: "2025-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Installing Bioconductor

```{r 1. Installing Bioconductor}
# if(!requireNamespace("BiocManager", quietly = TRUE))
#    installed.packages("BiocManager")BiocManager::install()

```

## 2. Install specific bioinformatics Packages

```{r 2. Install specific bioinformatics Packages}
# BiocManager::install("GenomicRanges")
# BiocManager::install("DESeq2")
# BiocManager::install("edgeR")

```

## 3. Installing Additional Useful Packages

### 3. A. Tidyverse - A collection of R packages for Data Science

```{r 3. A. Tidyverse - A collection of R packages for Data Science}
# install.packages("tidyverse")

```

### 3. B. Data Visualization

```{r 3. B. Data Visualization}
# install.packages("ggplot2")

```

### 3. C. Data Manipulation and Transformation 

```{r 3. C. Data Manipulation and Transformation}
# install.packages("dpyr")

```

### 3. D. Interactive Visualizations

```{r 3. D. Interactive Visualizations}
# install.packages("plotly")

```

## 4. Integrate version control with Git into your Bioinformatics workflow

[https://git-scm.com/](https://git-scm.com/)

```{r 4. Integrate version control with Git into your Bioinformatics workflow}
# [https://git-scm.com/](https://git-scm.com/)

```

## 5. Testing Your Environment

### 5. A. Loading Packages

```{r 5. A. Loading Packages}
library(GenomicRanges)
library(DESeq2)

```

### 5. B. Check Versions

```{r 5. B. Check Versions}
packageVersion("GenomicRanges")
packageVersion("DESeq2")

```

## 6. Session Info

```{r 6. Session Info}
devtools::session_info()
Sys.Date()

```




