---
title: "Chapter 8 - Gene Expression Analysis: RNA-Seq"
author: "Sasha Ajay Malkani"
date: "2026-01-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Gene Expression Analysis: RNA-Seq

As one of the cornerstone technologies in molecular biology, RNA sequencing 
(RNA-Seq) has revolutionized the way we analyze gene expression. RNA-Seq allows 
for the quantification of RNA levels across the transcriptome, providing insights
into gene regulation, alternative splicing, and even the discovery of novel 
transcripts. Given the complexity of RNA-Seq data sets, proper analysis is 
essential to extract meaningful biological insights. This chapter delves into 
the key steps involved in RNA-Seq data analysis using the R programming language,
a powerful tool for statistical computing and graphics.

### 1. A. Setting the Stage: Data Acquistion

The journey of RNA-Seq analysis begins with data aquisition. RNA-Seq experiments
can be performed in various organisms and conditions. The raw output typically 
consists of a series of FASTQ files, each corresponding to a sequencing run. 
These files contain both the sequence information and the quality scores for 
the reads. Before diving into analysis, it;s crucial to ensure data integrity 
and quality.

### 1. B. Quality Control

Quality Control (QC) is a vital step in RNA-Seq analysis. Tools such as FastQC 
and MultiQC can be used to visualize the quality of sequence reads. In R, the 
"Short Read" and "rtracklayer" packages can be employed to import and manipulate 
FASTQ files. Basic QC metrics include:

* Read length distribution

* Base quality scores

* Sequence duplication levels

* Adapter contamination

#### 1. B. I. R code example for quality control using FastQC:

```{r 1. B. I. R code example for quality control using FastQC}
# Load library
library(ShortRead)
library(dada2)

# Define the file path
fastq_file <- "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/SRR7508943.fastq/SRR7508943.fastq"

# Load the FASTQ file
fastq_data <- readFastq(fastq_file)

# Inspect the first few reads
head(fastq_data)

# Plot read quality
plotQualityProfile(fastq_file, n = 1013912)

```

### 1. C. Preprocessing: Trimming and Alignment

Following QC, pre-processing steps such as trimming and alignment are necessary.
Read trimming involves removing low-quality bases and adapter sequences, often 
done using tools like Trimmomatic or Cutadapt, which can be run outside of R. 
Once reads are trimmed, they must be aligned to reference genome.

#### 1. C. I. Alignment Tools

The R package "Rsubread" provides functionalities to perform read alignment 
using methods like subread, which is both fast and accurate. Once aligned, the 
output can be a BAM file containing the mapped reads.

R code for aligning reads:

```{r 1. C. I. Alignment Tools}
# Install package
# BiocManager::install("Rsubread")

# Install seqinr if not already installed
# install.packages("seqinr")

# Load Libraries
library(Rsubread)
library(seqinr)

# read.fasta() reads sequences from a FASTA-format file (including .fna)
# seqtype = "DNA" tells it the sequences are DNA
# as.string = TRUE keeps each sequence as a single string
# write.fasta() writes the sequences to a new FASTA file
# names(sequences) uses the original identifiers for FASTA headers

# Specify input and output file paths
input_file <- "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/ncbi_dataset/ncbi_dataset/data/GCA_000001735.2/GCA_000001735.2_TAIR10.1_genomic.fna"
output_file <- "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/GCA_000001735.2_TAIR10.1_genomic.fa"

# Read sequences from the .fna file
sequences <- read.fasta(file = input_file, seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

# Write sequences to a .fa file
write.fasta(sequences = sequences,
            names = names(sequences),
            file.out = output_file)

library(Biostrings)

# Read the original FASTA
Thale_fa <- readDNAStringSet("D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/GCA_000001735.2_TAIR10.1_genomic.fa")

# Write into a new FASTA with line length 60
writeXStringSet(Thale_fa, "Thale_genome_fixed.fasta", width=60)


# Build index from a reference genome FASTA file
# Example: Thale genome FASTA
buildindex(
  basename = "Thale_index", 
  reference = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale_genome_fixed.fasta"
)

# Aligning RNA-Seq reads 
align(
  index = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale_index",  # basename from buildindex()
  readfile1 = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/SRR7508943.fastq/SRR7508943.fastq",
  output_file = "Thale_aligned.bam"
)

```

### 1. D. Quantifying Gene Expression

Once the reads are aligned, the next step is quantifying gene expression levels. 
This can be achieved using several approaches, but one of the most popular is 
counting the number of reads mapping to each gene.

#### 1. D. I. Counting Reads

The "featureCounts" function in the Rsubread package is commonly used for counting 
reads that map to genomic features. This function facilitates the generation of a 
count matrix, where rows represent genes and columns correspond to samples

R code for counting reads:

```{r 1. D. I. Counting Reads}
# Load Library
library(Rsubread)

# Count reads mapped to genes using a GTF annotation
counts <- featureCounts(
  files = "Thale_aligned.bam",       # BAM file
  annot.ext = "genomic.gtf",         # GTF file
  isGTFAnnotationFile = TRUE,        # Tell featureCounts it's a GTF
  GTF.featureType = "exon",          # Count exons (common for RNA-seq)
  GTF.attrType = "gene_id",          # Group by gene_id
  useMetaFeatures = TRUE,            # Sum counts per gene
  isPairedEnd = FALSE                # Set to FALSE if single-end
)

# Summary of count object
summary(counts)

# Extract the count matrix
count_matrix <- counts$counts
head(count_matrix)
summary(count_matrix)

```

### 1. E. Differential Expression Analysis

With the count matrix in hand, researchers can conduct differential expression 
(DE) analysis to identify genes with statistically significant changes in 
expression between conditions.

#### 1. E. I. Using DESeq2

The "DESeq2" package in R is widely used for DE analysis. It incorporates 
normalization methods to account for variations in sequencing depth and other 
technical biases. In this section, we will focus on preparing the data, 
fitting the DE model, and extracting results.

R code for differential expression analysis:

Simulate a large RNA-seq count matrix with n = 1,013,912 counts in total, spread 
across genes and samples, so you can run DESeq2 as if it were real data.

We can do this by:
  
* Choosing a realistic number of genes (e.g., 20k for Arabidopsis).
* Generating counts from a negative binomial distribution (common for RNA-seq).
* Making sure the total sum of counts equals 1,013,912.
* Assigning two conditions (control vs treated) with replicates.

Here’s the R code to generate it:

```{r 1. E. I. Using DESeq2}
# Load library
library(DESeq2)

# Parameters
set.seed(42)  # reproducibility
n_genes <- 20000   # number of genes
n_samples <- 4     # 2 control, 2 treated
total_counts <- 1013912

# Simulate counts from a negative binomial distribution
raw_counts <- matrix(
  rnbinom(n_genes * n_samples, mu = 50, size = 1/0.2),  # mean=50, dispersion=0.2
  nrow = n_genes,
  ncol = n_samples
)

head(raw_counts)

# Scale counts so total sum matches exactly 1,013,912
scale_factor <- total_counts / sum(raw_counts)
scale_factor
counts_scaled <- round(raw_counts * scale_factor)
head(counts_scaled)

# Assign gene IDs
rownames(counts_scaled) <- paste0("Gene", seq_len(n_genes))
colnames(counts_scaled) <- c("control_rep1", "control_rep2", "treated_rep1", "treated_rep2")
head(counts_scaled)

# Create sample metadata
colData <- data.frame(
  condition = factor(c("control", "control", "treated", "treated")),
  replicate = c(1, 2, 1, 2)
)
rownames(colData) <- colnames(counts_scaled)


# Build DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = counts_scaled,
  colData = colData,
  design = ~ condition
)

head(dds)

# Run DESeq2
dds <- DESeq(dds)
dds
res <- results(dds)
res

# Check total counts
sum(counts(dds))  # should be 1013912

```

The output from the "results" function provides log fold changes, statistical 
significance (p-values), and other key metrics needed for biological 
interpretation.

### 1. F. Visualization of Results

Once differential expression analysis is complete, effective visualization is 
crucial for interpreting and communicating findings. R offers a variety of 
plotting options, including:

* Volcano plots to visualise DE results

* Heatmaps for showing expression profiles in clusters

* PCA (Principal Component Analysis) to explore sample relationships.

R code for generating a volcano plot:

```{r 1. F. Visualization of Results}
# Load library
library(ggplot2)

# Create a volcano plot
results_df <- as.data.frame(res)
head(results_df)

ggplot(results_df, aes(x = log2FoldChange, y = -log10(pvalue))) + 
  geom_point(alpha = 0.5) + theme_minimal() + 
  labs(title = "Volcano Plot", x = "-Log2 Fold Change P-value")

```

RNA-Seq has become an indispensable method for studying gene expression at a 
comprehensive level. This chapter provided an overview of the critical steps in 
RNA-Seq data analysis using R, from data acquistion and quality control to 
alignment, quantification, differential expression analysis and visualization. 
As researchers continue to explore the complexities of the transcriptome, tools 
like R and its rich ecosystem of packages will remain vital in translating raw 
sequencing data into insightful biological discoveries.

## 2. Aligning and Counting RNA-Seq Reads

RNA sequencing (RNA-Seq) has revolutionized our understanding of gene expression 
by allowing researchers to quantify transcript abundance and discover novel 
isoforms. However, the analysis of RNA-Seq datais a complex process that requires 
precise techniques for aligning reads to a reference genome or transcriptome 
and counting the number of reads that map to each gene. In this chapter, we 
guide you through the process of aligning and counting RNA-Seq reads using the 
R programming language and Bioconductor packages, providing you with a robust 
framework for analyzing your RNA-Seq data.

### 2. A. Prerequisites

Before you drive into RNA-Seq data analysis, it is crucial to have a working 
knowledge of R and some familiarity with genomic concepts such as sequences, 
genes and transcripts. Ensure that you have the following software  installed:

* __R (version 4.0 or higher)__

* __Bioconductor__ (installation instructions can be found on the Bioconductor 
website). We will make use of several packages which can be installed using 
Bioconductor:

```{r 2. A. Prerequisites}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.22")

# BiocManager::install(c("GenomicRanges", "Rsubread", "DESeq2", "edgeR"))

```

### 2. B. Understanding RNA-Seq Data

RNA-Seq data typically starts with raw sequencing reads, which can be obtained 
from high-throughput sequencing platforms such as Illumina. These reads must be 
processed and aligned to a reference geneome or transcriptome to facilitate 
downstream analysis. The primary steps in the RNA-Seq workflow include:

* __Quality Control__: Assessing the quality of raw sequencing data using tools 
like FastQC.

* __Alignment__: Mapping reads to the reference genome or transcriptome

* __Counting Reads__: Summarizing the number of reads that align to each gene.

* __Differential Expression Analysis__: Identifying genes that show significant 
changes in expression across conditions.

### 2. C. Quality Control of RNA-Seq Reads

Before alignment, it is essential to perform quality control on your raw RNA-Seq 
data. Tools like "FastQC" provide insights into sequencing quality, adapter 
content and read length distribution. In R, you can use the "fastqcr" package 
to visualize quality metrics.

```{r 2. C. Quality Control of RNA-Seq Reads}
# Install "fastqcr" package
# install.packages("fastqcr", repos = "http://cran.us.r-project.org")

# Load "fastqcr" package
library(fastqcr)

# Define the file path
fastq_file <- "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/SRR7508943.fastq/SRR7508943.fastq"

# Load the FASTQ file
fastq_data <- readFastq(fastq_file)

# Inspect the first few reads
head(fastq_data)

# Fastq function only works in command line (Linux) environment
# qc <- fastqc(fq.dir = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/SRR7508943.fastq/")
# Error in .check_if_unix() : Unix system (MAC OSX or Linux) required.

# Alternative use Short Read package
library(ShortRead)
fq <- readFastq(fastq_file)
qa <- qa(fastq_file)
report(qa, dest = "QA_Report")

```

### 2. D. Aligning Reads

Once quality control is complete, the next step is aligning the reads to the 
reference genome. R provides a handy interface with the "Rsubread" package, 
which includes the "align" function for this purpose.

#### 2. D. I. Indexing the Reference Genome

Before aligning, you need to create an index of the reference genome:

```{r 2. D. I. Indexing the Reference Genome}
# Build index from a reference genome FASTA file
# Example: Thale genome FASTA
# buildindex(
#   basename = "Thale_index", 
#   reference = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale_genome_fixed.fasta"
# )

```

#### 2. D. II. Aligning the Reads

Now, you can align your reads to the indexed reference genome:

```{r 2. D. II. Aligning the Reads}
# Aligning RNA-Seq reads 
# align(
#   index = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale_index",  # basename from buildindex()
#   readfile1 = "D:/Ajay Files/R Programming for Bioinformatics/Chapter 8/Thale Cress genome/SRR7508943.fastq/SRR7508943.fastq",
#   output_file = "Thale_aligned.bam"
# )

```

### 2. E. Counting Reads

After Alignment, the final step before differential expression analysis is 
counting how many reads to map to each gene. This is essential for constructing 
a count matrix for downstream analysis.

#### 2. E. I. Count Reads using "featureCounts"

"featureCounts", which is part of the "Rsubread" package, provides a simple way 
to count reads for annotated features like genes.

```{r 2. E. I. Count Reads using "featureCounts"}
# Count reads mapped to genes using a GTF annotation
# counts <- featureCounts(
#   files = "Thale_aligned.bam",       # BAM file
#   annot.ext = "genomic.gtf",         # GTF file
#   isGTFAnnotationFile = TRUE,        # Tell featureCounts it's a GTF
#   GTF.featureType = "exon",          # Count exons (common for RNA-seq)
#   GTF.attrType = "gene_id",          # Group by gene_id
#   nthreads = 4,                      # no. of. threads
#   useMetaFeatures = TRUE,            # Sum counts per gene
#   isPairedEnd = FALSE                # Set to FALSE if single-end
# )

# Summary of count object
# summary(counts)

```

#### 2. E. II. Organizing the Count Data

The output of "featureCounts" is a "SummarizedExperiment" object, which can be 
further manipulated to fit your analysis needs

```{r 2. E. II. Organizing the Count Data}
# Load Library
library(SummarizedExperiment)

# Convert count data to a matrix and set row and column names.
class(counts)
count_matrix_1 <- as.matrix(counts$counts)
rownames(count_matrix_1) <- counts$annotation$GeneID
colnames(count_matrix_1) <- counts$annotation$SampleID
head(count_matrix_1)
class(count_matrix_1)

```

We have walked through the essential steps for aligning and counting RNA-Seq 
reads using R. This process is foundational for any downstream analysis, 
facilitating differential expression studies and paving the way for various 
other bioinformatics analyses. As you become more familiar with these tools, 
you'll find that they offer great flexibility and power in handling RNA-Seq 
data.

## 3. Differential Gene Expression Analysis with DESeq2 and edgeR

This chapter focuses on two of the most popular tools for differential gene 
expression analysis: DESeq2 and edgeR. Both of these R packages are designed 
to perform statistical analysis on count data derived from RNA-seq experiments, 
and they offer robust methodologies for identifying differentially expressed 
genes (DEGs).

### 3. A. Overview of RNA-seq Data

RNA-seq generates vast amounts of data that represent the quantity of RNA in a 
sample at the time of sequencing. The resulting raw data-count matrices, where 
rows correspond to genes and columns represent samples - need pre-processing 
and normalization before differential expression analysis. Some common pre-
processing steps include filtering lowly expressed genes, normalizing for 
differences in library sizes, and removing batch effects if necessary. 

### 3. B. Introduction to DESeq2

#### 3. B. I. DESeq2 Overview

DESeq2 is an R package developed for analyzing count data from RNA-seq experiments. 
It employs a model based on the negative binomial distribution to account for the 
variability in gene expression levels. The workflow includes several key steps:

* __Input Data__: DESeq2 requires a matrix of raw counts and a design formula 
that describes the experimental conditions.

* __Normalization__: The package automatically normalizes data to correct for 
differences in library size by using size factors estimated from the counts.

* __Model Fitting__: DESeq2 fits a statistical model to the counts based on 
the design formula, estimating the dispersion of counts for each gene.

* __Differential Expression Testing__: Using the fitted model, DESeq2 performs
hypothesis testing to identify DEGs, controlling for false discovery rate (FDR)
via the Benjamini-Hochberg procedure.

#### 3. B. II. Key features of DESeq2

* __Robust Dispersion Estimates__: DESeq2 provides informed estimates of variance, 
which account for differences in biological variability across samples.

* __Visualization Tools__: It includes facilities for visualizing results through 
heatmaps, volcano plots, and principal component analysis (PCA).

* __Handling of Batch Effects__: The package allows users to include batch 
information in the model, mitigating the impact of unwanted variation on 
differential expression.

### 3. C. Introduction to edgeR

#### 3. C. I. edgeR Overview

edgeR is another widely used R package for analyzing RNA-seq count data, 
especially beneficial for experiments with small sample sizes. It also employs 
a negative binomial model for count data and utilizes empirical Bayes methods 
to enhance statistical power.

* __Input Data__: Similar to DESeq2, edgeR requires a count matrix and design 
matrix.

* __Data Normalization__: edgeR accounts for differences in sequencing depth 
and RNA composition using the trimmed mean of M-values (TMM) normalization 
method.

* __Dispersion Estimation__: It estimates gene-specific dispersions from the 
data and applies an empirical Bayes approch to improve estimation reliability.

* __Differential Expression Analysis__: The package uses likelihood ratio tests 
and exact tests to identify DEGs, providing a framework to perform multiple 
testing corrections.

#### 3. C. II. Key features of edgeR

* __Flexibility__: edgeR can handle complex experimental designs and is capable
of incorporating group and time-course comparisons.

* __Powerful Statistical Tests__: The package's robust statistical framework 
allows researchers to detect DEGs with high precision, particularly in low-count 
situations.

* __Visualization Capabilities__: edgeR offers multiple visualization options, 
including MA plots and heatmaps.

### 3. D. Comparative Analysis of DESeq2 and edgeR

While both DESeq2 and edgeR are powerful tools for differential gene expression 
analysis, they have different strategies, strengths and weaknesses.

#### 3. D. I. Similarities

* Both packages support the analysis of count data modelled by the negative 
binomial distribution.

* Both utilize empirical methods for estimating dispersion and controlling for 
FDR.

* Both offer extensive visualization capabilities to interpret results effectively.

#### 3. D. II. Differences

* __Normalization Methodology__: DESeq2 uses a median-of-ratios method for 
normalization, while edgeR employs the TMM normalization.

* __Models of Variability__: DESeq2 provides gene-specific variance estimates, 
while edgeR improves variance estimation through empirical Bayes methods, which 
might offer more stability in some contexts.

* __Ease of Use__: DESeq2 is often considered more straightforward for beginners 
due to its automated normalization and more guided workflow.

### 3. E. Case Study: Application of DESeq2 and edgeR

To illustrate the application of DESeq2 and edgeR, we present a hypothetical 
case study analyzing gene expression changes in cancer research. A researcher 
aims to identify DEGs between tumor and normal tissues using RNA-seq data.

#### 3. E. I. Step-by-Step Analysis

* __Data Pre-processing__: The raw count data is obtained and pre-processed, 
filtering low-expression genes and ensuring quality control.

* __Analysis with DESeq2__: 

The DESeqDataSet is created and normalized.

Differential expression is tested between tumor ans normal samples.

Results are visualized using PCA plots and heatmaps to highlight significant genes

* __Analysis with edgeR__: 

The DGEList object is contructed and normalized using TMM.

Likelihood ratio tests are performed to identify DEGs between the two groups.

Results are visualized using MA plots and volcano plots.

* __Comparison of Results__: The output from DESeq2 and edgeR is compared, with 
attention paid to overlapping DEG's, as well as unique findings from each analysis. 
The biological significance of these genes is discussed concerning relevant 
pathways in cancer.

### 3. F. Summary and Future developments

Differential gene expression analysis using DESeq2 and edgeR provides invaluable 
insights into genomic underpinnings of biological phenomena. Choosing between 
these two tools often depends on the specific dataset and experimental design. 
Regardless of the choice, proper understanding of their methodologies and 
careful application will yield meaningful interpretations, advancing our 
knowledge in genomics and precision medicine. As researchers continue to harness
the power of RNA-seq, DESeq2 and edgeR remain essential tools in the ongoing
exploration of gene expression dynamics.

With the continuous development of bioinformatics tools, new methodologies for 
RNA-seq data analysis will emerge, potentially integrating machine learning 
algorithms and enhanced visualization techniques. Future versions of DESeq2 and 
edgeR are likely to incorporate these advancements, streamlining workflows and 
improving accuracy in differential gene expression analysis. Researchers should 
stay updated with the latest releases and best practices to ensure the integrity 
and reproducibility of their findings.

## 4. Session Info

```{r 4. Session Info}
devtools::session_info()
Sys.Date()

```

